\documentclass[11pt]{article}

\usepackage{amsmath,amssymb}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{geometry,comment}
\usepackage{xcolor}
\geometry{margin=1in}
\usepackage{tikz}
\usetikzlibrary{decorations.pathreplacing}
\usepackage{tikz-cd}
\usepackage{pgfplots}
\usepackage[binary-units=true]{siunitx}
\usepackage{graphicx}
\usepackage[font=small]{caption}
\usepackage[most]{tcolorbox}

\newcommand{\rnote}[1]{{\color{red} {Ron: #1}}}

\title{SP1 zkVM Architecture}
\author{Succinct Labs\\[8pt]}

\input{macros.tex}
\date{\today}
\begin{document}

\maketitle

\section*{Overview}

This document describes the architecture of the \textbf{SP1 zkVM}
according to the W1 deliverable of the EF zkVM whitepaper
guidelines \cite{efZkvmGuidelines}.
SP1 proves execution of RISC-V guest programs using:

\begin{itemize}
\item \hyperref[sec:exec]{Execution of RV64IM programs} encoded as ELF binaries
\item \hyperref[sec:seg]{Sharded execution traces} with segmentation logic by trace area estimation.
\item Specialized computational units at the arithmetization layer called \hyperref[sec:core]{``chips''}, whose constraints
enforce correct execution of RISC-V instructions and cryptographic operations.
\item \hyperref[sec:bus]{Local bus interactions} (LogUp + GKR) for inter-chip communication within each shard.
\item \hyperref[sec:memory]{Global bus interactions} across shards (with septic curve accumulation).
\item A multilinear \hyperref[sec:proof-system]{proof system} (``Hypercube'') with BaseFold polynomial commitments over the KoalaBear field.
\item \hyperref[sec:proof-arch]{Recursive aggregation} via normalize / compose / shrink circuits
\end{itemize}

\section{Execution Model}\label{sec:exec}
We define how a program is executed.

\subsection{Program, Execution, and VM State}

A \textbf{program} $\PPP$ is a RISC-V ELF binary targeting the RV64IM
instruction set (64-bit base integer with multiply/divide extension).

The program $\PPP$ operates on the \textbf{VM state} $S$, defined as:
\begin{itemize}
\item $\pc$: 64-bit program counter
\item $\regs[0..31]$: 32 general-purpose registers (x0--x31), each 64 bits wide,
      with x0 hardwired to zero
\item $\mem[]$: byte-addressed writable memory
\end{itemize}
We separately denote the code of the program by $\code$. It is immutable and read-only, and is not part of the VM state.

A single instruction step of the program execution is defined as a transition from
one VM state to another according to the semantics of the instruction $\op$ retrieved from
$\code$ at the program counter $\pc_1$. This is denoted as follows:
\begin{equation}\label{eq:op}
S:=(\pc_1,\regs_1,\mem_1)\xrightarrow{\op:=\code[\pc_1]} S':=(\pc_2,\regs_2,\mem_2)
\end{equation}
For example, if $\op$ is an addition instruction, which adds the first two registers and stores in the third register, then the semantics of the step is defined as
\begin{align}
\pc_2&=\pc_1+4,\\
\regs_2&=\regs_1\text{ except that } \regs_2[2]:=\regs_1[0]+\regs_1[1],\\
\mem_2&=\mem_1.
\end{align}
We will use the notation $(\pc_1,\regs_1,\mem_1)\rightarrow  (\pc_2,\regs_2,\mem_2)$ to denote that $S$ transitions to $S'$ by executing instruction
 $\op:=\code[\pc_1]$.


An \textbf{execution trace} for program $\code$ applied to inputs $\III$ is a sequence of VM states

\[
\TTT:\quad S_0 \rightarrow S_1 \rightarrow \cdots \rightarrow S_T
\]
where each step corresponds to execution of one instruction.


A \textbf{zkVM proof} for $\PPP$ and initial state $S_0$ is a proof of the existence of an execution
trace $\TTT$ as above with $S_0$ as the initial state.

\paragraph{Supported Instructions.}
SP1 supports the full RV64IM instruction
set \cite{riscv}:
the base 64-bit integer instructions (I) and the integer
multiply/divide extension (M).

\paragraph{Execution Environment.}
Additionally, SP1 provides \emph{precompiles}---accelerated
implementations of expensive operations---invoked via the RISC-V
\texttt{ecall} instruction with a syscall code in register \texttt{t0}.

The precompiles include standard system-level ecalls---halt, I/O (write, hint read/len, commit), commit---
as well as the following accelerated cryptographic operations
(we include here only the precompiles relevant to the ethereum proving usecase):
\begin{itemize}
\item \textbf{Hash functions:} SHA-256 (extend + compress), Keccak-256
  permutation, Poseidon2.
\item \textbf{Elliptic curve operations:} point addition, doubling, and
  decompression for Secp256k1, Secp256r1, and BN-254.
\end{itemize}

\paragraph{Predicates.}
For our purpose the exact semantics of each instruction is not
important; we treat them as black boxes with predicates modeling
correctness.
For a given instruction $\op$, the predicate $\varphi_{\op}$ has the
semantics
\begin{multline}\label{eq:phiop}
\varphi_{op}(\pc_1,\regs_1,\mem_1,\pc_2,\regs_2,\mem_2)=\mathrm{TRUE} \;\Longleftrightarrow\;\\
 (\pc_1,\regs_1,\mem_1)\xrightarrow{\op} (\pc_2,\regs_2,\mem_2) \wedge \code[\pc_1]=op.
\end{multline}

\section{Segmentation}\label{sec:seg}
To facilitate efficient proving of large computations, SP1 splits the execution into ``shards''. We now explain how SP1 undertakes this splitting.

\subsection{Execution Shards}

After execution, the execution trace $\TTT$ is divided into
\emph{execution shards} $\GGG_1,\GGG_2,\ldots,\GGG_m$
(cf.\ Fig.~\ref{img:trace-seg}).
Execution shards are not fixed-size; instead, the executor splits the trace when
the estimated trace area of the current shard exceeds a
configurable threshold or more rarely if any one opcode is invoked more than a fixed number of times\footnote{The default element threshold is
$2^{28}+2^{27}$, and the maximum opcode count is $2^{22}$.}.
Each execution shard $\GGG_i$ corresponds to a contiguous subsequence of the
execution trace, from $S_{\mathrm{start}_i}$ to $S_{\mathrm{end}_i}$.

Sharding reduces the proof of the entire execution to proving the
validity of each execution shard and consistency between them.
The global consistency argument asserts:
$$
S_{\mathrm{end}_i} = S_{\mathrm{start}_{i+1}}\text{ for all }i\in[m-1],\quad S_{\mathrm{start}_1}=S_0,\quad S_{\mathrm{end}_m}=S_T.
$$

\subsection{Shard layout}\label{sec:core}
We now introduce the notion of a \emph{proving shard} (a distinct notion from \emph{execution} shard),
which we also abbreviate to ``shard'' in what follows.

Each shard is characterized by a collection of
chips, specialized ``zkVM computational units''.
Each chip is represented by a table, and typically corresponds to a RISC-V instruction,
a cryptographic operation, or some other fixed-size computation.
The columns corresponds to the operands relevant to the given operation, as well as auxiliary ``helper columns''.
The rows correspond to the different executions of the instruction/operation/computation in the given shard.
The chip is further associated with algebraic
constraints (AIRs) between its columns, which are designed to check all or part of the operation's
correctness. Constraints are proved via the zerocheck protocol, described below.

The remainder of a chip's correctness is checked via message passing.
Message-passing between chips within a shard is enabled by a \emph{local bus}; the matching of sent
and received messages is proved via the LogUp/GKR lookup argument \cite{DBLP:journals/iacr/PapiniH23}.
% a LogUp interaction argument (proved via GKR) \rnote{can we add a bibliography? happy to take care of it} that checks that all
% ``send'' and ``receive'' messages across chips balance within the shard. \rnote{worth mentioning that the bus is also used to enfore local constraints on the chip as well e.g. range checks}

All inter-shard communication passes through the global
accumulation bus, arithmetized as a \texttt{Global} table present in all shards (we describe the global
consistency argument in more detail below).
Each shard is responsible for computing and proving the correctness of $\sigma$, its contribution to the
global multiset equality accumulator.

Each shard proof attests to the relation
$$
\relation_{\mathrm{shard}}^{(C)} \;=\; \left\{
\left((\mathsf{pv},\sigma);\; \mathsf{trace}\right) \;\Bigg|\;
\begin{array}{l}
\text{All chip constraints in cluster $C$ are satisfied}\\
\text{All local-scope bus interactions balance (via LogUp)}\\
\sigma \text{ is the correct global cumulative sum}\\
\text{Public values $\mathsf{pv}$ are consistent with the trace}
\end{array}
\right\}
$$
where $C$ is the chip cluster determining the circuit, $\mathsf{pv}$ are the shard's public values
(boundary state, digests, etc.), and $\sigma$ is the shard's global cumulative sum\footnote{The proof system
for this circuit is named ``core'' in \href{https://github.com/ethereum/soundcalc/blob/main/soundcalc/zkvms/sp1/sp1.toml}{soundcalc}}.

Now, we can describe the relationship between execution shards and proving shards.
Each execution shard leads to at least one proving shard, but that proving shard may still need
to delegate some of the precompile correctness to another proving shard.
Moreover, global memory correctness requires the addition of proving shards that don't arise from
execution shards.
With this in mind, we describe the four types of chip clusters available for proving shards:

\begin{enumerate}
\item \textbf{Base core cluster} (\texttt{core}).
Contains all RISC-V instruction chips (ALU, memory load/store, control flow, etc.)
together with infrastructure chips: \texttt{Program} (instruction lookup),
\texttt{ByteLookup} and \texttt{RangeLookup} (preprocessed lookup tables),
\texttt{MemoryLocal} (local memory state),
\texttt{MemoryBump} and \texttt{StateBump} (timestamp management),
\texttt{SyscallCore}, \texttt{SyscallInstrs} (syscall dispatch),
and \texttt{Global} (global bus accumulator).

\item \textbf{Extended core clusters} (\texttt{core+}$P$).
Some precompiles are compact enough to be retained in the core cluster,
avoiding the overhead of a separate shard proof and allowing communication
with their calling context via the local lookup argument.
In particular, SHA hashing and Bn254 field operations are proved alongside RISC-V instructions.
Extended clusters are generated for subsets of size 0, 1, or all of these extensions,
plus one additional combination.

\item \textbf{Deferred precompile clusters} (\texttt{precompile-}$P$).
Precompiles with high arithmetization costs (in terms of number of columns
in the arithmetization) are proved in their own shards. Each such cluster contains
the precompile-specific chip(s) together with a base set of infrastructure chips:
\texttt{Program}, \texttt{ByteLookup}, \texttt{RangeLookup},
\texttt{SyscallPrecompile}, \texttt{MemoryLocal}, and \texttt{Global}.
They communicate with their RISC-V calling context
via the global lookup argument.
The deferred precompiles include more expensive elliptic curve operations (e.g. Secp256k1, Secp256r1, BN-254 add/double)
and the Keccak hash functions.

\item \textbf{Memory boundary cluster} (\texttt{mem-boundary}).
The global memory argument requires initialization and
finalization: sending the initial memory state into the global bus and receiving the final memory state
from the global bus.
The memory boundary cluster arithmetizes this process.
This cluster contains \texttt{MemoryGlobalInit}, \texttt{MemoryGlobalFinal},
\texttt{Global}, and the preprocessed chips (\texttt{Program}, \texttt{ByteLookup}, and \texttt{RangeLookup}).
\end{enumerate}

\begin{figure}
\centering
\resizebox{0.75\textwidth}{!}{%
\begin{tikzpicture}[y=-1cm]
  % Colors
  \definecolor{alucol}{RGB}{195,195,235}
  \definecolor{memcol}{RGB}{185,225,185}
  \definecolor{hashcol}{RGB}{205,205,205}
  \definecolor{poscol}{RGB}{245,215,175}
  \definecolor{buscol}{RGB}{255,195,115}
  \definecolor{globalcol}{RGB}{180,220,255}
  %
  \tikzset{
    chip/.style={draw, thick, rounded corners=4pt, minimum width=2.2cm, minimum height=0.7cm,
                 font=\footnotesize\sffamily\bfseries, align=center},
    busbar/.style={draw, thick, rounded corners=3pt,
                 minimum height=0.5cm, font=\footnotesize\sffamily\bfseries},
  }
  %
  % === SHARD BOX ===
  \draw[semithick, rounded corners=6pt] (0, 0) rectangle (11.5, 5.2);
  \node[font=\small\sffamily\bfseries, anchor=north west] at (0.15, 0.15) {Shard};
  %
  % === TOP ROW OF CHIPS ===
  \node[chip, fill=alucol!70] (alu) at (1.8, 1.2) {\texttt{Add}};
  \node[chip, fill=memcol!70] (mem) at (4.5, 1.2) {\texttt{MemoryLocal}};
  \node[chip, fill=alucol!70] (ctrl) at (7.2, 1.2) {\texttt{JALR}};
  \node[chip, fill=hashcol!50] (prog) at (9.9, 1.2) {\texttt{Program}};
  %
  % === LOCAL BUS ===
  \node[busbar, fill=buscol, minimum width=10.5cm] (lbus) at (5.75, 2.7) {Local Bus};
  %
  % === BOTTOM ROW OF CHIPS ===
  \node[chip, fill=hashcol!70] (keccak) at (1.8, 4.0) {\texttt{Keccak}};
  \node[chip, fill=memcol!70] (memloc) at (4.5, 4.0) {\texttt{SHA256Compress}};
  \node[chip, fill=cyan!30] (range) at (7.2, 4.0) {\texttt{RangeCheck}};
  \node[chip, fill=globalcol] (global) at (9.9, 4.0) {\texttt{Global}};
  %
  % === ARROWS: TOP CHIPS <-> LOCAL BUS ===
  \draw[<->, thick] (alu.south) -- (alu.south |- lbus.north);
  \draw[<->, thick] (mem.south) -- (mem.south |- lbus.north);
  \draw[<->, thick] (ctrl.south) -- (ctrl.south |- lbus.north);
  \draw[<->, thick] (prog.south) -- (prog.south |- lbus.north);
  %
  % === ARROWS: BOTTOM CHIPS <-> LOCAL BUS ===
  \draw[<->, thick] (keccak.north) -- (keccak.north |- lbus.south);
  \draw[<->, thick] (memloc.north) -- (memloc.north |- lbus.south);
  \draw[<->, thick] (range.north) -- (range.north |- lbus.south);
  \draw[<->, thick] (global.north) -- (global.north |- lbus.south);
  %
  % === GLOBAL BUS ===
  \node[busbar, fill=globalcol!70, minimum width=11.5cm] (gbus) at (5.75, 6.4) {Global Bus};
  %
  % === ARROW: SHARD → GLOBAL BUS ===
  \draw[<->, very thick] (5.75, 5.2) -- (gbus.north);
  %
\end{tikzpicture}%
}
\caption{Shard structure: chips communicate via the local bus within a shard, and each shard interacts with the global bus.}\label{img:segments}
\end{figure}

\section{Correct execution of a program}
In this section, we define via predicates what it means that a program is correctly executed. We will use these predicates in Section~\ref{sec:proof-arch} to define the relations that the zkVM proves.

\subsection{Predicates for individual operations}

Each operation is treated as a black box with its
own predicate $\varphi_{\op}$ (cf.~\eqref{eq:phiop}).
Throughout this section, subscripts $1$ and $2$ denote states before and after an operation.
We write $S_1:=(\pc_1,\regs_1,\mem_1)$ and
$S_2:=(\pc_2,\regs_2,\mem_2)$ for the states
before and after the operation.

We group operations as follows:
\begin{itemize}
\item \textbf{ALU operations.}
  Integer arithmetic (ADD, SUB, MUL, DIV, REM, and their variants),
  bitwise operations (AND, OR, XOR), shifts (SLL, SRL, SRA),
  and comparisons (SLT, SLTU).
  Each has a predicate $\pred{\op}(S_1,S_2)$.
\item \textbf{Memory operations.}
  Loads (LB, LH, LW, LD and unsigned variants) and
  stores (SB, SH, SW, SD).
  Load: $\pred{load}(S_1,S_2)$ --- reads memory into a register.
  Store: $\pred{store}(S_1,S_2)$ --- writes a register value to memory.
\item \textbf{Control flow.}
  Branches (BEQ, BNE, BLT, BGE, BLTU, BGEU), jumps (JAL, JALR),
  and upper-immediate (LUI, AUIPC).
\item \textbf{Precompiles.}
  Invoked via \texttt{ecall}; each precompile $P$ has a predicate
  $\pred{P}(S_1,S_2)$ covering all state changes
  (register and memory) produced by the syscall.
\end{itemize}

\subsection{Correct execution of a single step}

A single step executes the instruction at the current program counter.
The step predicate selects the matching operation from the program code:
\begin{equation*}\label{eq:step}
\varphi_{\step}(S_1,S_2):= \varphi_{\code[\pc_1]}(S_1,S_2)
\end{equation*}
where $S_1$ and $S_2$ are the states before and after the step.

\subsection{Chip interactions and the bus}\label{sec:bus}

Chips communicate through a \emph{bus}: a collection of
typed messages that chips \emph{send} (produce) and \emph{receive}
(consume).
Consistency is enforced by requiring that, for each interaction kind,
the multiset of sent messages equals the multiset of received messages.
A structured message sent/received by all rows of a table is termed an \emph{interaction}

% SP1 implements this multiset equality via a \textbf{LogUp argument} (proved via GKR):
% each chip contributes terms $+1/(\alpha - m_i)$ for sends and
% $-1/(\alpha - m_j)$ for receives, where $\alpha$ is a random
% challenge and $m_i, m_j$ are message fingerprints.
% The sum must equal zero, which holds if and only if sends and receives
% match as multisets \rnote{(whp over $\alpha$)}.

Each interaction is declared with one of two \emph{scopes}:
\begin{itemize}
\item \textbf{Local} interactions must balance within the shard.
This is enforced by the LogUp/GKR argument inside the shard proof.
\item \textbf{Global} interactions need not balance within a single
shard.
Instead, each shard computes its net contribution
$\sigma\in E(\mathbb{F})$, a point on a
elliptic curve defined on a degree-7 extension of the base field
and outputs it as a public value.
The contributions are accumulated across all shards during recursion (or core proof verification),
and the final completeness check asserts
$\sum_i \sigma_i = \mathcal{O}$ (the identity point).
\end{itemize}

The global scope is used for interactions that cross shard boundaries,
most notably memory consistency and precompile delegation to separate
shards.

\subsection{Memory consistency}\label{sec:memory}

In SP1, memory consistency is enforced via the memory-checking argument of Blum et al. \cite{Blum}.
Namely, for each memory read or write, one adds an ($addr$, $value$, $timestamp$) tuple to a write set $WS$
and a read set $RS$ (for reads the same values are written to both).
For either read or write, the $addr$ value added to both sets needs to be the same, and the timestamp
added to $WS$ needs to be later than the one added to $RS$.
To initialize memory, one adds the initial value of all touched memory to $WS$; similarly,
one adds the final value of all touched addresses to $RS$.
Memory consistency is then equivalent to set equality $WS=RS$, together with strict monotonicity of
the timestamps.
This argument is used at both the local and global scopes, using the corresponding buses for proving
equality of sets.

\paragraph{Local/Intra-Shard Memory Consistency.}
Within a shard, local memory interactions (via the \texttt{MemoryLocal}
chip) handle the read/write operations that occur entirely within that
shard, using the local LogUp argument for consistency.
It is necessary to pass initial and final memory states into the global bus for all addresses touched
during the shard's execution. However, even if an address is touched multiple times within the shard,
it is necessary only to pass its initial and final values into the global bus.

\paragraph{Global/Cross-Shard Memory Consistency.}
At the beginning of execution, a dedicated chip sends each initial memory cell
$(addr, value, timestamp)$ into the global bus.
At the end of execution, another chip receives
the final memory state from the global bus.
As mentioned above, each shard also sends its initial and final memory states into the global bus.
The final global completeness check ($\sigma = \mathcal{O}$) ensures that
every memory cell written is eventually read with the correct value,
and that the initial and final memory states are consistent with the
full execution.


\section{Recursion Architecture}\label{sec:proof-arch}
In the previous two sections, we described how a proof of correct RISC-V execution is produced by
dividing the computation into shards, arithmetizing shard correctness via constraints and lookups,
and finally gluing together the shard proofs via a global bus consistency argument.
The number of shards scales roughly with the cycle count of the execution; to produce constant-size
proofs it is therefore necessary to introduce some form of recursion or proof aggregation.
In this section, we describe the approach taken by SP1 to address this problem.

It is useful when discussing recursion to set some terminology
to avoid the confusion caused by the fact that a recursion \emph{prover} is proving the \emph{verifier} of another
stage of the pipeline.
First, by a \textbf{recursion program} or \textbf{recursion circuit}, we mean the program or arithmetic cirucit that verifies
one or more SNARK proofs and also optionally checks some aggregation logic.
SNARK proofs can typically be verified in the circuit model of computation, so the term ``circuit'' here is suitable.
Given a recursion circuit $\mathcal C$, when we say a $\mathcal C$ proof, we mean a proof of the correct
execution of $\mathcal C$.
Often, the inputs to a recursion circuit are $\mathcal{C}'$ proofs for some other circuit $\mathcal C'$;
we advise the reader to be aware that the term $\mathcal C$ proof refers to the \emph{output} of proving
$\mathcal C$.

\paragraph{Proving recursion circuits.}  The recursion prover, like the core prover, arithmetizes
the computation into a ``chips and bus'' architecture; unlike the core prover, there is no need for
cross-shard communication (a single circuit proof fits comfortably into one proving shard), and therefore
no global bus. The chips used for proving recursion circuits include a \texttt{Poseidon2} chip for proving
Poseidon2 hashes, KoalaBear field arithmetic tables, and other operations performed in bulk by the Hypercube
proof system's verifier.


\subsection{Topology}

The recursion proof pipeline consists of multiple circuit types, summarized in the table below.
~\footnote{The proof parameters for \texttt{normalize} and \texttt{compose} are the same, so in \href{https://github.com/ethereum/soundcalc/blob/main/soundcalc/zkvms/sp1/sp1.toml}{soundcalc}, they are listed under the name ``compress''.
The proof parameters for \texttt{shrink} are listed in soundcalc under ``shrink''.
}

\begin{center}
\begin{tabular}{lll}
\hline
\textbf{Circuit} & \textbf{Purpose} & \textbf{Relation}\\
\hline
% \texttt{core} (many shapes) & Core shard proof & $\relation_{\mathrm{shard}}^{(C)}$ \\
\texttt{normalize}          & 1-to-1 shape normalization & $\relation_{\mathrm{norm}}$ \\
\texttt{compose}            & $N$-to-1 aggregation ($2\!\leq\! N\!\leq\! 4$) & $\relation_{\mathrm{compose}}$ \\
\texttt{shrink}             & Final proof size reduction & $\relation_{\mathrm{shrink}}$ \\
\hline
\end{tabular}
\end{center}

In more detail:

\begin{itemize}
% \item \texttt{core}:
% Proves the correct execution of one shard.
% The circuit shape varies with the chip cluster type and trace
% dimensions (Sec.~\ref{sec:core}).
\item \texttt{normalize}:
A 1-to-1 circuit that verifies a single core shard proof. A \texttt{normalize} proof
is always of a fixed shape (which we call \texttt{compress\_shape}), irrespective of
the input proof's shape (Sec.~\ref{sec:normalize}).
\item \texttt{compose}:
An $N$-to-1 recursion circuit ($2\leq N\leq 4$) that verifies multiple
proofs of shape \texttt{compress shape} and performs some aggregation logic.
A \texttt{compose} proof is always a single proof of shape \texttt{compress\_shape}
(Sec.~\ref{sec:compose}).
\item \texttt{shrink}:
A 1-to-1 circuit that takes the final complete \texttt{compose}
proof and re-proves it with smaller FRI parameters, reducing
the proof size in preparation for the outer proof system
(Sec.~\ref{sec:shrink}).
\end{itemize}

We visualize the recursion topology in Fig.~\ref{fig:topo}.
At the leaves, \texttt{core} shard proofs are each normalized to \texttt{compress\_shape} by a
\texttt{normalize} circuit.
The resulting proofs are then aggregated by a tree of \texttt{compose} circuits
(fan-in up to~4).
We refer to the \texttt{normalize} and \texttt{compose} layers
collectively as the ``compress'' stage.
The compress stage terminates when all proofs have been
aggregated into a single complete \texttt{compose} proof.
Finally, this proof is passed to the \texttt{shrink} circuit, which verifies
the complete \texttt{compose} proof.
The proving system for the \texttt{shrink} circuit produces smaller proofs
because it 1) does not normalize its proof to fit \texttt{compress\_shape} and 2) it uses
more favorable FRI parameters.


\begin{figure}
\centering
\begin{tikzpicture}[
  box/.style={draw, rounded corners=3pt, minimum height=0.65cm,
              inner sep=3pt, font=\scriptsize\sffamily, align=center},
  core/.style={box, fill=yellow!20, minimum width=1.0cm},
  norm/.style={box, fill=green!15, minimum width=1.0cm},
  comp/.style={box, fill=blue!12, minimum width=1.2cm},
  shrk/.style={box, fill=orange!15, minimum width=1.2cm},
  arr/.style={draw, thin, -stealth},
]
% --- Core shards (row 0) ---
% LHS: 4 cores, evenly spaced
\node[core] (c1) at (0,    0) {core};
\node[core] (c2) at (1.3,  0) {core};
\node[core] (c3) at (2.6,  0) {core};
\node[core] (c4) at (3.9,  0) {core};
% RHS: 3 cores, same 1.3 spacing, centered under RHS compose
\node[core] (c5) at (5.85, 0) {core};
\node[core] (c6) at (7.15, 0) {core};
\node[core] (c7) at (8.45, 0) {core};
% --- Normalize (row 1) ---
\node[norm] (n1) at (0,    1.15) {norm};
\node[norm] (n2) at (1.3,  1.15) {norm};
\node[norm] (n3) at (2.6,  1.15) {norm};
\node[norm] (n4) at (3.9,  1.15) {norm};
\node[norm] (n5) at (5.85, 1.15) {norm};
\node[norm] (n6) at (7.15, 1.15) {norm};
\node[norm] (n7) at (8.45, 1.15) {norm};
\foreach \i in {1,...,7} { \draw[arr] (c\i) -- (n\i); }
% --- Compose layer 1 (row 2): fan-in 4 (LHS), fan-in 3 (RHS) ---
\node[comp] (p1) at (1.95, 2.3) {compose};
\node[comp] (p2) at (7.15, 2.3) {compose};
\draw[arr] (n1) -- (p1);
\draw[arr] (n2) -- (p1);
\draw[arr] (n3) -- (p1);
\draw[arr] (n4) -- (p1);
\draw[arr] (n5) -- (p2);
\draw[arr] (n6) -- (p2);
\draw[arr] (n7) -- (p2);
% --- Compose layer 2 (row 3) ---
\node[comp] (p3) at (4.55, 3.45) {compose};
\draw[arr] (p1) -- (p3);
\draw[arr] (p2) -- (p3);
% --- Shrink (row 4) ---
\node[shrk] (sh) at (4.55, 4.6) {shrink};
\draw[arr] (p3) -- (sh);
% --- "compress" brace on the left (encompasses normalize + compose layers) ---
\draw[decorate, decoration={brace, amplitude=6pt}]
  (-1.3, 0.83) -- (-1.3, 3.78)
  node[midway, left=8pt, font=\scriptsize\sffamily, align=center] {compress};
\end{tikzpicture}
\caption{\label{fig:topo} Recursion topology.
Colors:
\colorbox{yellow!20}{\textsf{\scriptsize core}}\,,
\colorbox{green!15}{\textsf{\scriptsize normalize}}\,,
\colorbox{blue!12}{\textsf{\scriptsize compose}}\,,
\colorbox{orange!15}{\textsf{\scriptsize shrink}}\,.
The tree is illustrative; the actual fan-in and depth depend on the
number of shards.}
\end{figure}

\paragraph{Verifying key Merkle tree.}
Each recursion circuit type may have multiple \emph{shapes}---distinct
circuit configurations that arise from different chip clusters and
trace dimensions for the circuit inputs.
Each shape leads to a distinct circuit with its own verifying key $\vk$.
These keys are organized into a Merkle tree whose root
$\mathsf{vk\_root}$ is used to authenticate recursion proofs.
The Merkle proof guarantees that the corresponding recursion proof was of the correct
execution of \emph{one} of the recursion circuits and not some other, unrelated circuit.

The shapes are enumerated as follows:
\begin{itemize}
\item \textbf{Normalize} (the vast majority of circuits):
one shape per combination of chip cluster, preprocessed and main
stacked BaseFold column counts, and padding column counts.
\item \textbf{Compose}: one shape per arity ($N=2,3,4$).
\item \textbf{Shrink}: a single shape.
\end{itemize}

The \texttt{compose} and \texttt{shrink} programs enforce VK authentication:
they takes Merkle membership proofs for each input proof's verifying key
as part of their witness, verify them against the Merkle root, and
check that every input proof's $\mathsf{rpv}.\mathsf{vk\_root}$
equals this root.
We write $\varphi_{\mathrm{auth}}(\vk, \mathsf{mkl}, \mathsf{vk\_root})$
for the predicate asserting that $\vk$ is a valid leaf of the Merkle tree
with root $\mathsf{vk\_root}$, verified via membership proof $\mathsf{mkl}$.
The \texttt{normalize} program does not perform VK membership
checks itself; a \emph{consumer} of a \texttt{normalize} proof (typically the \texttt{compose} circuit, also possibly the Rust verifier
of the proof) does, however, need to check the correctness of the \texttt{normalize} verifying key.

Note that, because of the 1-1 normalization, there truly is exactly one \texttt{compose} circuit for each
arity.
If not for the normalization, the number of arity $N$ \texttt{compose} circuits would be roughly $C^N$,
where $C$ is the number of core shard proof shapes.



\subsection{\texttt{normalize} circuit}\label{sec:normalize}
The \texttt{normalize} circuit proves the correct verification of a single
core shard proof. Its output is a proof of a fixed shape (\texttt{compress\_shape}),
irrespective of the input shard proof's shape (which varies with the chip cluster).
The circuit takes as witness a shard proof $\pi$, the verifying key $\vk$,
a completeness flag $\mathsf{is\_complete}\in\{0,1\}$,
and a VK Merkle root $\mathsf{vk\_root}$.

The circuit performs two categories of work:
\begin{itemize}
\item \textbf{Shard proof verification.}
A Fiat--Shamir challenger is initialized by observing the verifying key,
and the core shard proof is verified against this challenger
(polynomial commitment openings and constraint satisfaction).

\item \textbf{Aggregation checks.}
Outside the shard verifier, the circuit enforces cross-shard and global correctness:
\begin{itemize}
\item \emph{First-shard PC validation.}
If the shard is the first execution shard,
the circuit asserts that $\mathsf{pc\_start}$ matches the verifying key's start PC.

\item \emph{Global cumulative sum accumulation.}
The output global cumulative sum $\sigma$ is computed by adding
the verifying key's $\mathsf{initial\_global\_cumulative\_sum}$ (if this is the first shard,
otherwise $\mathcal{O}$) to the shard's own $\mathsf{global\_cumulative\_sum}$.

\item \emph{Public values propagation.}
The shard's public values (boundary PCs, timestamps, memory addresses,
digests, exit codes, etc.)\ are propagated into a \texttt{RecursionPublicValues} struct
together with the VK digest and VK Merkle root.
A Poseidon-2 digest of these values is committed as the proof's public output.

\item \emph{Completeness conditions.}
When $\mathsf{is\_complete}=1$, the proof is expected to attest to a
full program execution rather than a single shard, so final boundary
conditions are checked --- including that the program has halted,
the first shard was included,
all ``previous'' boundary values are at their initial state,
and the global cumulative sum equals $\mathcal{O}$
(the identity point on the septic curve).
\end{itemize}
\end{itemize}

Formally, let $\Pi_{\mathrm{shard}}^{(C)}$ be the SNARK for
$\relation_{\mathrm{shard}}^{(C)}$.
Write $\mathsf{pv}$ for the shard's public values
(including its global cumulative sum $\sigma_{\mathrm{shard}}$)
extracted from the proof,
and $\mathsf{rpv}$ for the recursion public values output by the circuit.

$$\relation_{\mathrm{norm}}=\left\lbrace\begin{array}{c|c}
  \mathsf{rpv}
  ;\;
  (\pi,\,\vk,\,C)
  &
  \begin{array}{l}
  \Pi_{\mathrm{shard}}^{(C)}.\mathsf{Verify}(\mathsf{pv},\,\pi)=1\\[4pt]
  \mathsf{rpv}.\mathsf{contains\_first\_shard}=1
    \;\Longrightarrow\;
    \mathsf{pv}.\mathsf{pc\_start}=\vk.\mathsf{pc\_start}\\[4pt]
  \mathsf{rpv}.\sigma \;=\;
    \sigma_{\mathrm{init}} + \mathsf{pv}.\sigma_{\mathrm{shard}},\quad
    \sigma_{\mathrm{init}}:=
    \begin{cases}
      \vk.\sigma_0 & \text{if first shard}\\
      \mathcal{O} & \text{otherwise}
    \end{cases}\\[4pt]
  \varphi_{\mathrm{output}}(\mathsf{rpv};\,\mathsf{pv},\,\vk)\\[4pt]
  \mathsf{rpv}.\mathsf{is\_complete}=1
    \;\Longrightarrow\;
    \varphi_{\mathrm{complete}}(\mathsf{rpv})
  \end{array}
\end{array}\right\rbrace$$
where $\varphi_{\mathrm{complete}}$ asserts the final boundary conditions
described above (program halted, first shard included, initial boundary
values zero, $\sigma=\mathcal{O}$),
and $\varphi_{\mathrm{output}}(\mathsf{rpv};\,\cdot)$ asserts that the output
recursion public values $\mathsf{rpv}$ are correctly derived from the given
inputs, including a valid Poseidon-2 digest commitment.

\subsection{\texttt{compose} circuit}\label{sec:compose}

The \texttt{compose} circuit is an $N$-to-1 recursion circuit ($1\leq N\leq 4$)
that consumes $N$ proofs --- each of which may be a \texttt{normalize}
or \texttt{compose} proof --- and produces one proof of
the fixed shape \texttt{compress\_shape}.
Its purpose is to aggregate a batch of proofs that cover
contiguous portions of the execution into a single proof.

The circuit performs two categories of work:
\begin{itemize}
\item \textbf{Proof verification.}
Each input proof $\pi_i$ is verified against its own verifying key $\vk_i$.
The circuit checks that every proof's $\mathsf{vk\_root}$ matches
a single witnessed value, and that each proof's recursion public values
have a valid digest.

\item \textbf{Continuity and aggregation checks.}
The circuit enforces that the $N$ proofs form a contiguous chain:
\begin{itemize}
\item \emph{Sequential continuity.}
For each consecutive pair of proofs,
the ``last'' boundary values of proof~$i$
(PC, timestamp, memory addresses, page indices,
exit code, syscall flags, etc.)
must equal the ``previous'' boundary values of proof~$i{+}1$.
\item \emph{Program identity.}
The $\mathsf{sp1\_vk\_digest}$ and $\mathsf{proof\_nonce}$ are
asserted to be identical across all $N$ proofs.
\item \emph{Accumulation.}
The global cumulative sums $\sigma_i$ from all proofs are summed
(via elliptic curve addition on the septic curve).
The shard counts and $\mathsf{contains\_first\_shard}$ flags
are accumulated.
\item \emph{Completeness conditions.}
When $\mathsf{is\_complete}=1$,
the same final boundary conditions as in the \texttt{normalize} circuit
are checked via $\varphi_{\mathrm{complete}}$.
\end{itemize}
\end{itemize}

Formally, let $\Pi_{\mathrm{comp}}$ denote the SNARK for \texttt{compress\_shape} proofs
(used by \texttt{normalize} and \texttt{compose} alike).
Write $\mathsf{rpv}_i$ for the recursion public values of the $i$-th input proof
($1\leq i\leq N$, $1\leq N\leq 4$),
and $\mathsf{rpv}$ for the aggregated output.

$$\relation_{\mathrm{compose}}=\left\lbrace\begin{array}{c|c}
  \mathsf{rpv}
  ;\;
  (\{\vk_i,\pi_i,\mathsf{mkl}_i\}_{i=1}^{N},\;\mathsf{vk\_root})
  &
  \begin{array}{l}
  \forall\, i:\;
    \Pi_{\mathrm{comp}}.\mathsf{Verify}(\mathsf{rpv}_i,\,\pi_i)=1
    \;\wedge\;
    \mathsf{rpv}_i.\mathsf{vk\_root}=\mathsf{vk\_root}\\[4pt]
  \forall\, i:\;
    \varphi_{\mathrm{auth}}(\vk_i,\,\mathsf{mkl}_i,\,\mathsf{vk\_root})\\[4pt]
  \forall\, i<N:\;
    \mathsf{rpv}_i.\mathsf{last\_*}
    = \mathsf{rpv}_{i+1}.\mathsf{prev\_*}\\[4pt]
  \forall\, i:\;
    \mathsf{rpv}_i.\mathsf{sp1\_vk\_digest}
    = \mathsf{rpv}_1.\mathsf{sp1\_vk\_digest}\\[4pt]
  \mathsf{rpv}.\sigma \;=\;
    \textstyle\sum_{i=1}^{N} \mathsf{rpv}_i.\sigma\\[4pt]
  \varphi_{\mathrm{output}}(\mathsf{rpv};\,\{\mathsf{rpv}_i\}_{i=1}^{N})\\[4pt]
  \mathsf{rpv}.\mathsf{is\_complete}=1
    \;\Longrightarrow\;
    \varphi_{\mathrm{complete}}(\mathsf{rpv})
  \end{array}
\end{array}\right\rbrace$$

\noindent
Here ``$\mathsf{last\_*}=\mathsf{prev\_*}$'' is shorthand for the continuity of all
chained boundary values
(PC, timestamp, memory init/finalize addresses, page indices,
exit code, and syscall flags).

\subsection{\texttt{shrink} circuit}\label{sec:shrink}

The \texttt{shrink} circuit is a 1-to-1 recursion circuit that takes the
final complete \texttt{compose} proof and re-proves it with reduced FRI
parameters.
Its purpose is to reduce the proof size: the compress stage uses
FRI parameters sized for a log-trace-area of~$27$, while the shrink
circuit uses parameters for a log-trace-area of~$25$, yielding a
significantly smaller proof that is cheaper to verify in the subsequent
outer proof system (e.g., Groth16).
The shrink circuit has a single fixed shape.

The circuit takes as witness
a compress-stage proof $\pi$,
the verifying key $\vk$ of the input proof,
and a VK Merkle membership proof $\mathsf{mkl}$.

The circuit performs the following work:
\begin{itemize}
\item \textbf{VK Merkle verification.}
The input proof's verifying key is checked against the
VK Merkle tree, ensuring it belongs to the set of
permitted recursion circuits.

\item \textbf{Proof verification.}
The compress-stage proof $\pi$ is verified against its verifying key
$\vk$ using the compress verifier
(the same verifier used inside \texttt{compose}).

\item \textbf{Completeness enforcement.}
The circuit asserts that the input proof is marked
$\mathsf{is\_complete}=1$, meaning it attests to a full program
execution---the program has halted, all shards have been aggregated,
the global cumulative sum equals $\mathcal{O}$, and all boundary
conditions are satisfied.

\item \textbf{Output construction.}
The output recursion public values are derived from the input proof's
public values.
A Poseidon-2 digest of the output public values is committed as the
shrink proof's public output.
\end{itemize}

Formally, let $\Pi_{\mathrm{comp}}$ denote the SNARK for
\texttt{compress\_shape} proofs (as in the \texttt{compose} circuit).
Write $\mathsf{rpv}'$ for the recursion public values extracted from
the input proof and $\mathsf{rpv}$ for the output.

$$\relation_{\mathrm{shrink}}=\left\lbrace\begin{array}{c|c}
  \mathsf{rpv}
  ;\;
  (\vk,\,\pi,\,\mathsf{mkl})
  &
  \begin{array}{l}
  \Pi_{\mathrm{comp}}.\mathsf{Verify}(\mathsf{rpv}',\,\pi)=1\\[3pt]
  \varphi_{\mathrm{auth}}(\vk,\,\mathsf{mkl},\,\mathsf{rpv}'.\mathsf{vk\_root})\\[3pt]
  \mathsf{rpv}'.\mathsf{is\_complete}=1\\[3pt]
  \varphi_{\mathrm{complete}}(\mathsf{rpv}')\\[3pt]
  \varphi_{\mathrm{output}}(\mathsf{rpv};\,\mathsf{rpv}')
  \end{array}
\end{array}\right\rbrace$$

\noindent
Because shrink enforces completeness, its output serves as the
definitive attestation that the entire program execution is valid.
The shrink proof is then passed to the outer proof system for final
verification.

\section{Proof system instantiations}\label{sec:proof-system}

In the previous sections we described how to arithmetize correct RISC-V execution
by decomposing it into a collection of tables of field elements, each with algebraic constraints
on its columns, and
allowing tables to \emph{send} and \emph{receive} messages with multiplicity through
a collection of interaction kinds.
All tables in a chip cluster are zero-padded to a common height $\mathcal{N}_{row}=2^{n_{row}}$;
for core circuits $n_{row}=22$.
Throughout this section, let $W$ denote the total number of columns across all tables in the shard.

The proof of each shard follows a four-step protocol:
\begin{enumerate}
\item \textbf{Commit.}
      Treat the unpadded trace data as a single dense multilinear polynomial~$Q$
      and commit to it via (interleaved) Reed--Solomon encoding and Merkle hashing
      (Sec.~\ref{sec:trace-commit}).
\item \textbf{LogUp + GKR.}
      Prove that all local-scope interactions balance
      (sends $=$ receives as multisets).
      The output is a collection of $W$ evaluation claims on individual trace columns,
      all at a single point $\eta$
      (Sec.~\ref{sec:logup-gkr}).
\item \textbf{Zerocheck.}
      Prove that every algebraic constraint is satisfied on
      all unpadded rows.
      Batch the constraint zerocheck with the LogUp+GKR opening claims
      into a single sumcheck whose output is a new collection of $W$ evaluation claims at a
      point~$\eta'$
      (Sec.~\ref{sec:zerocheck}).
\item \textbf{PCS opening.}
      Reduce the $W$ column claims at~$\eta'$ to a single
      claim on~$Q$ via the jagged PCS (Sec.~\ref{sec:jagged}),
      then prove the claim on~$Q$ via BaseFold (Sec.~\ref{sec:dense-pcs}).
\end{enumerate}

Because every table is padded to the same height $\mathcal{N}_{row}$,
the verifier can be described as an arithmetic circuit that depends on the chip
cluster and the total unpadded trace area, but is \emph{independent} of the individual
tables' unpadded heights.
The total number of distinct verifier circuits is therefore bounded by $\approx 2^{18}$,
which is small enough that all circuits can be pre-built;
no VM is needed in the recursion layer.

If the prover actually materialized the full $W\times\mathcal{N}_{row}$ padded trace,
the data volume would be enormous:
$W\approx 2^{12}$ columns at 4~bytes each gives
$4\times 2^{12}\times 2^{22}=2^{36}$ bytes $= 64\,$GB.
The jagged PCS (Sec.~\ref{sec:jagged}) ensures that both prover and verifier costs
depend only on the size of the \emph{unpadded} data.


We now proceed to describe the above steps in more detail.

\medskip\noindent
\textbf{Notation.}
Whenever a variable appears in \texttt{typewriter font},
it refers to the corresponding parameter in the
\texttt{sp1.toml} configuration file in the \texttt{soundcalc} repository.

%% ================================================================
\subsection{Field and Fiat--Shamir challenger}\label{sec:field}

\paragraph{Base field.}
SP1 operates over the \textbf{KoalaBear} prime field
$\mathbb{F}_p$ with $p=2^{31}-2^{24}+1$.


\paragraph{Extension field.}
All verifier randomness is sampled over the degree-4 extension
\[
  \mathbb{F}_{p^4} \;=\; \mathbb{F}_p[x]\big/(x^4-3),
\]
a binomial extension.
We write $\mathbb{F}$ for the base field and $\mathbb{E}$ for the extension
when the distinction matters (\texttt{field = "KoalaBear\^{}4"} in \texttt{sp1.toml}).

\paragraph{Fiat--Shamir challenger.}
Interactive challenges are derived non-interactively via a
\textbf{Poseidon-2 duplex sponge}.
The sponge state has width~16 base-field elements;
a digest consists of 8 base-field elements
($8\times 31 = 248$ bits; cf.\ \texttt{hash\_size\_bits = 248}).

\paragraph{Proof-of-work grinding.}
At three points in the protocol the prover must perform proof-of-work
grinding before the verifier supplies the next challenge:
\begin{itemize}
\item Before batching BaseFold evaluation claims
      \texttt{grinding\_batching\_phase} bits.
\item Before the FRI query phase in BaseFold:
      \texttt{grinding\_query\_phase} bits.
\item Before sampling the randomness used for LogUp fingerprinting (\texttt{grinding\_bits\_lookup}).
\end{itemize}
These grinding steps raise the cost of a forging attempt
without increasing verifier complexity.

%% ================================================================
\subsection{Trace commitment}\label{sec:trace-commit}

All unpadded table data is flattened into a single array.
This array is divided into chunks of size
$\texttt{dense\_length}$ (a power of two);
each chunk becomes one ``column'' of a dense multilinear polynomial~$Q$.
There are \texttt{dense\_batch} such chunks in total, so
$Q$ is a multilinear polynomial in
$\log_2(\texttt{dense\_length})+\lceil\log_2(\texttt{dense\_batch})\rceil$ variables.
Alternatively, think of $Q$ as a column-major matrix of height $\texttt{dense\_length}$
and width \texttt{dense\_batch}.

The prover commits to $Q$ as follows:
\begin{enumerate}
\item Interpret each chunk as the coefficients of a univariate polynomial of
      degree $=\texttt{dense\_length}-1$.
\item Reed--Solomon encode each chunk at rate $\rho=1/\texttt{blowup\_factor}$,
      producing codewords of length
      $\texttt{dense\_length}\times\texttt{blowup\_factor}$.
\item Hash the encoded columns into a single Merkle tree (with the entries of each ;
      the Merkle root constitutes the trace commitment. The $i$-leaf of the Merkle tree has size
      \texttt{dense\_batch}, and consists of the $i$-th entry of each codeword.
\end{enumerate}
The prover also reports the unpadded height $h_i$ of each table.
Together with the commitment, these heights fully determine the trace.

%% ================================================================
\subsection{LogUp + GKR}\label{sec:logup-gkr}

The goal of this step is to prove that all \emph{local-scope} interactions
balance: for each interaction kind, the multiset of sent messages equals
the multiset of received messages.
(Recall from Sec.~\ref{sec:bus} that
global interactions are handled by the septic curve accumulator.)

\paragraph{LogUp equation.}
A chip cluster defines a maximum number \texttt{num\_columns\_S} of interactions found in the chips,
and a maximum message size \texttt{num\_lookups\_M} for all chips
Each message~$\mathbf{v} = (v_1,\ldots, v_{\texttt{num\_lookups\_M}})$ sent through the bus with index $BI$ is mapped to a field element via a \emph{multilinear fingerprint}:
\[
  \mathrm{fp}(\mathbf{v}) \;=\; \mathrm{eq}_{\beta}(0)\cdot BI + \sum_{i=1}^{\texttt{num\_lookups\_M}-1} \mathrm{eq}_{\mathbf{\beta}}(i)\, v_i
  \qquad (\texttt{multilinear\_fingerprint = true}),
\]
where $\mathbf{\beta}\in \mathbb{E}^{\lceil{\log_2()\texttt{num\_lookups\_M})}\rceil}$ is randomness
sampled for the LogUp protocol.
For a given interaction, let $t$ denote the vector $\{\mathrm{fp}(\mathbf{v}_x)\}_{x\in \{0,1\}^{n_{row}}}$, where $\mathbf{v}_x$ is the message
passed from row $x\in \{0,1\}^{n_{row}}$.
Similarly, let $m(x)$ denote the multiplicity of the message at row $x$.
The verifier samples $\alpha\in \mathbb{E}$ and the prover computes and sends
\begin{equation}\label{eq:logup}
  \kappa:= \sum_{x\in\{0,1\}^{n_{row}}} \frac{m(x)}{\alpha - t(x)}.
\end{equation}
The verifier sums the $\kappa$ values across the interactions in the chip cluster, and rejects a proof
if the sum is non-zero (the consistency of bus sends and receives is equivalent to this sum being zero).

\paragraph{GKR reduction.} The above fractional sums can be described as the outputs of a circuit of
depth $n_{row}$; SP1 proves the correctness of this circuit evaluation via the GKR interactive proof \cite{DBLP:journals/iacr/PapiniH23}.

\paragraph{Output.}
After $n_{row}$ layers the GKR protocol terminates with evaluation claims on the multilinear extensions
$\hat{m}(\eta)$ and $\hat{t}(\eta)$.
The original vectors $m$ and $t$ are affine-linear combinations of the original columns of the chip.
The prover provides column openings of the chip at~$\eta$, and the verifier checks that the proper affine-linear combinations
of the chips match the claimed values for $\hat{m}(\eta), \hat{t}(\eta)$.

%% ================================================================
\subsection{Zerocheck}\label{sec:zerocheck}

The goal of this step is to prove that every algebraic constraint is
satisfied on the unpadded rows of every table, while simultaneously
proving the column evaluation claims produced by LogUp+GKR.

\paragraph{Setup.}
The chip cluster has $k$ tables.
Table~$i$ has $w_i$ columns
$p_{i,1},\ldots,p_{i,w_i}$, unpadded height~$h_i$,
and $c_i$ constraints $C_{i,1},\ldots,C_{i,c_i}$, each of total degree at
most \texttt{air\_max\_degree}$\,=3$.

\paragraph{Zerocheck sumcheck.}
The constraint and GKR opening claims are batched (via random challenges)
into a single sumcheck over $\{0,1\}^{n_{row}}$.
After $n_{row}$ rounds, the sumcheck reduces to an evaluation claim
on the batched polynomial at a random point~$\eta'$.
This claim involves the constraint polynomials~$C_{i,j}$ evaluated at
the columns of each table at~$\eta'$.
Since each $C_{i,j}$ is an explicit low-degree polynomial,
the verifier can check the claim given the individual column values
$p_{i,\ell}(\eta')$.
The prover therefore supplies these values, yielding
$W$ column evaluation claims at~$\eta'$.

%% ================================================================
\subsection{Jagged PCS}\label{sec:jagged}

The goal of this step is to reduce the $W$ column evaluation claims at~$\eta'$
to a single claim on the dense polynomial~$Q$ committed in
Sec.~\ref{sec:trace-commit}.

\paragraph{Column $\leftrightarrow$ dense relationship.}
Each padded column $p_\ell$ of height $\mathcal{N}_{row}$ is a multilinear
polynomial in $n_{row}$ variables.
Its unpadded entries occupy a contiguous slice of the dense array underlying~$Q$.
The jagged PCS~\cite{DBLP:journals/iacr/HemoJRRR25} bridges the two representations without the prover ever materializing
the padded columns.
The protocol can be described as a sumcheck over a product of multilinears, one of which is $Q$ and
the other of which can be efficiently evaluated.
We use the jagged assist (as described in the above-mentioned reference) to further reduce verifier work in the evaluation
of the latter polynomial.

\paragraph{Output.}
The jagged PCS reduces the $W$ column claims to a single claim
$Q(z)=v$ on the dense polynomial.

%% ================================================================
\subsection{Dense PCS (BaseFold)}\label{sec:dense-pcs}

The final step proves the evaluation claim $Q(z)=v$ using
stacked \textbf{BaseFold} \cite{DBLP:conf/crypto/ZeilbergerCF24}, the multilinear variant of FRI.

\paragraph{Protocol.}
BaseFold combines the standard FRI proximity test with a sumcheck
that establishes the multilinear evaluation of the committed oracle.
Recall that $Q$ is described as a table of height \texttt{dense\_length}
and width \texttt{dense\_batch}.
Let $n:=\lceil\log_2(\texttt{dense\_batch})\rceil$, $m:=\log_2(\texttt{dense\_length})$.
Decompose $z=(z_{stack}, z_{col})$, where $z_{stack}\in \mathbb{E}^m,\quad z_{col}\in \mathbb{E}^n$.
\begin{enumerate}
\item \textbf{Columnwise claims.} The prover sends multilinear evaluation claims of the columns of $Q$
at $z_{stack}$.
\item \textbf{Batching step.} The verifier samples randomness $\mathbf{\tau}\in \mathbb{E}^n$.
The prover takes the random linear combination of the columns of $Q$ via the coefficients $\hat{\mathrm{eq}}_{\mathbf{\tau}}(\cdot)$.
The prover and verifier take the same linear combination of the column evaluation claims.
In the remaining steps, prover and verifier engage in the BaseFold protocol for the random linear combination.
\item \textbf{Folding rounds.}
      In each round the prover sends a sumcheck message and folds the current polynomial by a factor of~$2$
      (as specified by \texttt{fri\_folding\_factors}),
      using the next sumcheck challenge as the folding randomness.
      The folded oracle is committed via Reed-Solomon encoding and a fresh Merkle tree.
\item \textbf{Early stop.}
      Folding stops when length of the R-S encoded oracle reaches
      \texttt{fri\_early\_stop\_degree}.
      The prover sends the remaining coefficients in the clear.
\item \textbf{Query phase.}
      The verifier samples \texttt{num\_queries} query positions.
      For each query the prover opens the corresponding leaf in every
      folding-round Merkle tree.
      Proof-of-work grinding of \texttt{grinding\_query\_phase} bits
      precedes this phase.
\end{enumerate}

\bibliographystyle{alpha}
\bibliography{references}

\end{document}
